Draft

### 2017-08-17 Thesis/Draft.tex
add description of experiments

@@

@@

\begin{longtable}[]{@{}lrrrrrrrrrr@{}}
\caption{Count rates for ten products for both training and test
sample.}\tabularnewline
\toprule
& V64 & V1 & V159 & V119 & V70 & V114 & V189 & V139 & V118 &
V148\tabularnewline
\midrule
\endfirsthead
\toprule
& V64 & V1 & V159 & V119 & V70 & V114 & V189 & V139 & V118 &
V148\tabularnewline
\midrule
\endhead
Training & 0.001 & 0.013 & 0.027 & 0.053 & 0.059 & 0.071 & 0.071 & 0.102
& 0.156 & 0.158\tabularnewline
Test & 0.000 & 0.011 & 0.022 & 0.055 & 0.058 & 0.064 & 0.060 & 0.115 &
0.150 & 0.130\tabularnewline
TrainingMinusTau & -0.033 & -0.021 & -0.006 & 0.019 & 0.026 & 0.038 &
0.038 & 0.068 & 0.123 & 0.125\tabularnewline
\bottomrule
\end{longtable}

@@

\begin{longtable}[]{@{}lrrrrrrrrrr@{}}
\caption{Count rates for ten products for both training and test
sample.}\tabularnewline
\toprule
& V129 & V32 & V76 & V103 & V149 & V104 & V110 & V150 & V112 &
V154\tabularnewline
\midrule
\endfirsthead
\toprule
& V129 & V32 & V76 & V103 & V149 & V104 & V110 & V150 & V112 &
V154\tabularnewline
\midrule
\endhead
Training & 0.000 & 0.000 & 0.012 & 0.044 & 0.049 & 0.053 & 0.055 & 0.095
& 0.123 & 0.199\tabularnewline
Test & 0.013 & 0.000 & 0.011 & 0.040 & 0.049 & 0.063 & 0.060 & 0.114 &
0.137 & 0.202\tabularnewline
TrainingMinusTau & -0.083 & -0.083 & -0.071 & -0.039 & -0.035 & -0.030 &
-0.028 & 0.011 & 0.039 & 0.116\tabularnewline
\bottomrule
\end{longtable}

@@

>>>>

\subsubsection{Discussion}\label{discussion-1}

At the beginning of this thesis, we set out to find an algorithm that
improves upon the Anytime Parameter Free algorithm introduced in
Locatelli et al. (2016) for the thresholding bandit problem in two
directions. First, we wanted to make the algorithm perform better in
problems that are not well approximated as sub-Gaussian; that is, in
particular a problem with Bernoulli arms with small means and low
threshold. Unsurprisingly, this meant that we need to find a solution
that takes more into account than just the sample mean. With the Simple
Likelihood Ratio strategy, we propose an algorithm that fulfills that
goal at the cost of being distribution specific.

The Simple Likelihood Ratio strategy requires the knowledge of the class
of distribution upfront, just as other algorithms require parameters.
While this is likely to be a negligiable problem in the case of
\(\{0,1\}\) observations which can be modeled reasonably by Bernoulli
distributions if we are only interested in estimating the mean, it can
be more difficult to know the true class of distribution upfront when
the random variables can take values in \(\mathbb{R}\). We have shown
however, that the SLR strategy can be adapted to all univariate
distributions in the exponential family of distributions. The algorithm
even offers a natural adaptation to the problem of normal distributions
with unknown mean and unknown variance.

Furthermore, the Simple Likelihood Ratio strategy reduces to the Anytime
Parameter Free strategy in the case where arms are normally distributed
with equal variance that is known upfront. In that case we have as index
(compare example 3 above):

\[
B_i^{SLR}(t) = T_i(t)\frac{(\hat{\mu}_i(t) - \tau)^2}{2\sigma^2}.
\] In order to decide at time \(t\) which arm to draw next, we
necessarily compare the indices of arm \(i\) and arm \(j\). Assume
w.l.o.g. that \(B_i^{SLR}(t) < B_j^{SLR}(t)\). Then:

\[
B_i^{SLR}(t) < B_j^{SLR}(t) \Leftrightarrow 
T_i(t)\frac{(\hat{\mu}_i(t) - \tau)^2}{2\sigma^2} < T_j(t)\frac{(\hat{\mu}_j(t) - \tau)^2}{2\sigma^2} 
\Leftrightarrow 
\sqrt{T_i(t)}\sqrt{(\hat{\mu}_i(t) - \tau)^2} < \sqrt{T_j(t)}\sqrt{(\hat{\mu}_j(t) - \tau)^2}
\Leftrightarrow 
B_i^{APT}(t) < B_j^{APT}(t)
\]

@@

\subsection{Synthetic Data}\label{synthetic-data}

In order to evaluate the empirical performance of the proposed
algorithms, we run the algorithm on synthetic data as is usual. We
compare its performance on this data not only against the APT algorithm,
but also against the variance-based algorithms. Thus we also compare the
variance-based algorithms for the first time directly against each other
instead of just relative to the APT algorithm.

Each of the experiments is repeated for 5000 iterations. Depending on
the experiment, the observations are generated randomly by some
distribution of the exponential family of distributions. As we were
originally concerned with improving upon APT with respect to
non-sub-Gaussian distributions, the first experiments focus on Bernoulli
distributions, and especially those with small means. We then go on to
show with experiments on Gaussian, Exponential, and Poisson data that
the SLR algorithm is able to improve upon APT given that we know the
true kind of distribution.

\subsubsection{\texorpdfstring{Experiment 1: Bernoulli with
\(\tau = 0.5\)}{Experiment 1: Bernoulli with \textbackslash{}tau = 0.5}}\label{experiment-1-bernoulli-with-tau-0.5}

The first experiment is equal to experiment 3 in Locatelli et al.
(2016). We compare ten Bernoulli arms that move away from the threshold
at \(\tau = 0.5\) geometrically with \(\mu_{1:4} = 0.4-0.2^{1:4}\),
\(\mu_5 = 0.45\), \(\mu_6 = 0.55\), and
\(\mu_{7:10} = 0.6 + 0.2^{5-1:4}\). We run the standard APT algorithm
with \(\epsilon = 0.1\), as well as SLR with \(\epsilon = 0.1\). We
compare those additionally to Augmented-UCB with the standard parameter
choice of \(\rho = 1/3\) as used by its authors in Mukherjee et al.
(2017). We run the parameter-free version of the Empirical Variance
Guided algorithm without any feedback delay, thus making it directly
comparable to the other algorithms. The Bayes-UCB algorithm takes as
prior a Beta distribution with parameters \(\alpha_0 = \tau\) and
\(\beta_0 = 1 - \tau\), thus leading to a mean of the prior distribution
at \(\tau\) while keeping the prior as weak as possible. Lastly, we run
the uniform sampling strategy as the naive benchmark.

\begin{longtable}[]{@{}rrrrrrrrrr@{}}
\caption{Mean of arms in simulation 1.}\tabularnewline
\toprule
V1 & V2 & V3 & V4 & V5 & V6 & V7 & V8 & V9 & V10\tabularnewline
\midrule
\endfirsthead
\toprule
V1 & V2 & V3 & V4 & V5 & V6 & V7 & V8 & V9 & V10\tabularnewline
\midrule
\endhead
0.2 & 0.36 & 0.392 & 0.398 & 0.45 & 0.55 & 0.602 & 0.608 & 0.64 &
0.8\tabularnewline
\bottomrule
\end{longtable}

This first experiment gives an experiment where all algorithms should
perform well. The problem is not very difficult, and due to the
threshold at \(\tau = 0.5\), the Bernoulli distribution can be perfectly
approximated by a Gaussian. This thus is the perfect environment for the
APT algorithm.

As the next figure shows, essentially all methods converge rapidly to an
error rate of zero. Except for the uniform method that clearly lacks
behind. We also observe that the variance based algorithms (Augmented
UCB and EVT) lag behind at first. Augmented UCB is able to adjust its
sampling scheme after 1000 iterations and rapdily improves to the level
of APT. EVT takes about 250 samples more to start a quick descend to the
performance of the other algorithms. All of this, however, happens below
an error rate of 0.01. While the SLR algorithm improves upon APT, the
improvement is a slight one, and not very large in terms of budget
difference. This underlines that the scenario at hand is the ideal one
for the APT algorithm.

\includegraphics{Draft_files/figure-latex/simulation1_results-1.pdf}

\subsubsection{\texorpdfstring{Experiment 2: Bernoulli with
\(\tau = 0.05\)}{Experiment 2: Bernoulli with \textbackslash{}tau = 0.05}}\label{experiment-2-bernoulli-with-tau-0.05}

In the second experiment, we compare the same algorithms as in the first
experiment. Again, we classify 10 arms, but this time against a
threshold of \(\tau = 0.05\) and without an epsilon interval,
\(\epsilon = 0\). Consequently, we choose the distributions for the arms
to be much smaller too, and arrange them symmetrically around the
threshold. While the distributions are bounded and thus still
sub-Gaussian, the Normal approximation should be poor in comparison to
the previous case. We expect an improvement for the SLR algorithm
compared to APT as it takes the Bernoulli characteristics into account.
For example, even though the arms are symmetrically distributed around
the threshold in terms of absolute distance, they are not symmetrically
around the threshold in terms of their Kullback-Leibler divergence given
the threshold is unequal 0.5. Thus we expect the SLR algorithm to not
pull them equally. Furthermore, we expect an improvement for the
variance-based algorithms, as the variances are now smaller than in
experiment 1, which should at least benefit the EVT algorithm given its
theoretical analysis.

The figure below shows the performance of all algorithms. We see that
the problem appears to be significantly more difficult in general. The
variance based algorithms perform maybe surprisingly bad; EVT even worse
than uniform sampling nearly until the horizon, while Augmented UCB
takes a long time to adjust away from uniform sampling. Both algorithms
perform worse than APT after 7000 rounds. One potential reason for this
might be the fact that the index used by the EVT algorithm reduces to
\(B_i^{EVT}(t) = \sqrt{T_i(t)} \cdot \sqrt{|\hat{\mu}_i(t) - \tau|}\) if
\(\hat{\sigma}_i^2(t) = 0\), which, given Bernoulli arms with small
means, could be the case for long runs of the algorithm. In contrast,
the index of the APT algorithm is
\(B_i^{APT}(t) = \sqrt{T_i(t)} \cdot |\hat{\mu}_i(t) - \tau|\).

One interesting observation, that we will observe in all experiments
with synthetic Bernoulli arms, is the superior performance of the
Bayes-UCB algorithm adapted from Kaufmann and Kalyanakrishnan (2013)
which was originally designed for cumulative regret problems.
Nonetheless, the algorithm will again and again proof to be the best
performing.

While the Bayes-UCB performance is impressive, we observe that also the
Simple Likelihood Ratio strategy improves considerably upon the
alternative strategies. SLR reaches the average empirical error of 0.1
within the budget, and the improvement is exponential within the entire
budget; in contrast to APT which appears to improve linearly in the
observations. At the end of the budget, the difference in the empirical
error is nearly 10\% between the two strategies.

In order to assess how the SLR strategy improves upon the APT algorithm,
we can use the fact that we have run the experiment for 5000 iterations.
We thus can look at how often each arm was pulled by a strategy at any
given round. Across the 5000 iterations, the algorithm should pull at
any given iteration the arm most often that it thinks is the most
difficult to classify. In the figure below, we plot the resulting curves
where the value at an iteration is the 10-round moving average to remove
some unimportant variation.

We see that both APT and SLR correctly single out the two arms that are
much closer to the threshold than the remaining arms and allocate most
of the budget to them. However, while SLR keeps the two arms constant at
two different levels (about 1900 for arm 5 and about 2500 for arm 6),
APT at first pulls arm 5 much more than arm 6. Then, the budget
allocated to the two arms appears to converge to about 2000. Given that
the two arms are symmetric around the threshold, it makes sense that APT
would allocate the same exploration effort to both of them. And given
that the variation of arm 5 away from the threshold is limited from
below by zero, it is much easier for arm 6 to have an empirical mean
larger than it's true mean, and thus to have a larger index than arm 5.
Consequently, APT would prefer to sample arm 5. This, however, is wrong
in terms of Kullback-Leibler divergence. Measured by the divergence, arm
6 is closer to the threshold than arm 5, thus harder to classify, and
consequently preferred by the SLR algorithm. Furthermore, the SLR
algorithm appears to quickly find the correct share that it has to
allocate to the two arms. In contrast to APT which continues to adjust
the shares across the entire budget.

Additionally, we can observe that the SLR strategy starts to drop arms
much quicker than APT, and it drops arm 1 to 3 first, which again
underlines the effect of the asymmetric Kullback-Leibler divergence. If
we consider arm 3 and arm 8, we have \(d(0.01||0.05) \approx 0.025\)
while \(d(0.09||0.05) = 0.014\). The divergence is much smaller for arm
8. The asymmetry is even more dramatic for arm 1 and arm 10 with
\(d(0.001||0.05) \approx 0.046\) and \(d(0.099||0.05) \approx 0.020\);
in terms of the divergence, the distance between arm 1 and the threshold
is twice as far as between arm 10 and the threshold. APT considers them
to be equally distant from the threshold.

\begin{longtable}[]{@{}rrrrrrrrrr@{}}
\caption{Mean of arms in simulation 2 classified against threshold of
0.05.}\tabularnewline
\toprule
V1 & V2 & V3 & V4 & V5 & V6 & V7 & V8 & V9 & V10\tabularnewline
\midrule
\endfirsthead
\toprule
V1 & V2 & V3 & V4 & V5 & V6 & V7 & V8 & V9 & V10\tabularnewline
\midrule
\endhead
0.001 & 0.005 & 0.01 & 0.015 & 0.04 & 0.06 & 0.085 & 0.09 & 0.095 &
0.099\tabularnewline
\bottomrule
\end{longtable}

\includegraphics{Draft_files/figure-latex/simulation2_results-1.pdf}

\subsubsection{\texorpdfstring{Experiment 3: Bernoulli with
\(\tau = 0.01\)}{Experiment 3: Bernoulli with \textbackslash{}tau = 0.01}}\label{experiment-3-bernoulli-with-tau-0.01}

In the third experiment, we lower the threshold further to
\(\tau = 0.1\) and arrange the arms on a \(\log_{10}\) scale around it.
Again, there is no epsilon interval. We add one additional arm, thus now
classifying eleven arms. We run the same arms as previously. Notice that
the experiment horizon is now \(10000\) rounds instead of \(7000\)
rounds as in the previous experiment. We test the same algorithms used
in the previous two experiments.

If we look at the performance after 7000 rounds, we see that the problem
appears to be slightly more difficult compared to the previous one. We
see that up until this point the variance-based strategies cannot
improve upon APT. Also, SLR has not yet crossed the 10\% empirical
error. However, we observe that in the following rounds EVT decreases at
an exponential rate and thus crosses the error of APT and continues to
improve upon it. At the end of the budget, the improvement of SLR
compared to the other algorithms is again very large, and likely to
increase. However, Bayes-UCB is again the best performing algorithm and
is improving at a faster rate than SLR. As in the previous experiment,
Bayes-UCB and SLR show significant improvements on this Bernoulli
problem with very small means and threshold compared to APT in
particular.

\begin{longtable}[]{@{}rrrrrrrrrrr@{}}
\caption{Mean of arms in simulation 2 compared to a threshold of
0.01.}\tabularnewline
\toprule
V1 & V2 & V3 & V4 & V5 & V6 & V7 & V8 & V9 & V10 & V11\tabularnewline
\midrule
\endfirsthead
\toprule
V1 & V2 & V3 & V4 & V5 & V6 & V7 & V8 & V9 & V10 & V11\tabularnewline
\midrule
\endhead
1e-04 & 3e-04 & 6e-04 & 0.001 & 0.0018 & 0.0032 & 0.0056 & 0.0141 &
0.0178 & 0.0316 & 0.1\tabularnewline
\bottomrule
\end{longtable}

\includegraphics{Draft_files/figure-latex/simulation3_results-1.pdf}

\subsection{\texorpdfstring{Experiment 4: Exponential Distribution with
\(\tau = 1\)}{Experiment 4: Exponential Distribution with \textbackslash{}tau = 1}}\label{experiment-4-exponential-distribution-with-tau-1}

\includegraphics{Draft_files/figure-latex/simulation4_results-1.pdf}

\subsection{\texorpdfstring{Experiment 5: Poisson Distribution with
\(\tau = 1\)}{Experiment 5: Poisson Distribution with \textbackslash{}tau = 1}}\label{experiment-5-poisson-distribution-with-tau-1}

\includegraphics{Draft_files/figure-latex/simulation5_results-1.pdf}

@@

\begin{longtable}[]{@{}lrrrrrrrrrr@{}}
\caption{Count rates for ten products for both training and test
sample.}\tabularnewline
\toprule
& V1 & V159 & V64 & V139 & V189 & V70 & V114 & V148 & V118 &
V119\tabularnewline
\midrule
\endfirsthead
\toprule
& V1 & V159 & V64 & V139 & V189 & V70 & V114 & V148 & V118 &
V119\tabularnewline
\midrule
\endhead
Training & 0.001 & 0.013 & 0.027 & 0.053 & 0.059 & 0.071 & 0.071 & 0.102
& 0.156 & 0.158\tabularnewline
Test & 0.000 & 0.011 & 0.022 & 0.055 & 0.058 & 0.064 & 0.060 & 0.115 &
0.150 & 0.130\tabularnewline
TrainingMinusTau & -0.033 & -0.021 & -0.006 & 0.019 & 0.026 & 0.038 &
0.038 & 0.068 & 0.123 & 0.125\tabularnewline
\bottomrule
\end{longtable}

@@

\begin{longtable}[]{@{}lrrrrrrrrrr@{}}
\caption{Count rates for ten products for both training and test
sample.}\tabularnewline
\toprule
& V112 & V129 & V104 & V76 & V103 & V110 & V150 & V149 & V32 &
V154\tabularnewline
\midrule
\endfirsthead
\toprule
& V112 & V129 & V104 & V76 & V103 & V110 & V150 & V149 & V32 &
V154\tabularnewline
\midrule
\endhead
Training & 0.000 & 0.000 & 0.012 & 0.044 & 0.049 & 0.053 & 0.055 & 0.095
& 0.123 & 0.199\tabularnewline
Test & 0.013 & 0.000 & 0.011 & 0.040 & 0.049 & 0.063 & 0.060 & 0.114 &
0.137 & 0.202\tabularnewline
TrainingMinusTau & -0.083 & -0.083 & -0.071 & -0.039 & -0.035 & -0.030 &
-0.028 & 0.011 & 0.039 & 0.116\tabularnewline
\bottomrule
\end{longtable}

@@

\includegraphics{Draft_files/figure-latex/unnamed-chunk-8-1.pdf}

We can try to understand why some strategies fail, while showing
significant improvements on synthetic data. Given that we run 5000
iterations of the experiment, we can assess in how many of these
iterations a given arm was pulled at a given round. Comparing these
sampling shares across algorithms may give us hints to as to why
algorithms perform a certain way. Given the setup with 5000 iterations
and 10 arms in the current experiment, we would expect for the uniform
strategy on average 500 pulls for each arm in every round. Given this
benchmark, we can now look at how the other strategies split their
budget. In the figures below, we have plotted not the absolute pulls per
round, but have applied a 10-round moving average to remove some of the
fluctuation in the plotted lines.

We observe that the patterns across algorithms are quite similar, while
the patterns are very different across data sets. Given that arm
\(V149\) is clearly the most difficult arm to classify due to its
closeness to the threshold \(\tau\), all algorithms focus a major share
of their budget on this arm. The speed at which they start to do so,
however, differs. Perhaps surprisingly, the best performing Bayes-UCB
algorithm increases its focus on arm \(V149\) the slowest, while in many
iterations spending budget on exploring other arms. Surprisingly because
Bayes-UCB is designed for the cumulative regret problem. Thus one would
expect it to explore \emph{less} than other algorithms. But its
performance does not seem to be easily explained by the lessened focus
on the most difficult arm; else we would expect the APT algorithm to
perform second best. However, both SLR and EVT perform better than APT
while converging more quickly to a higher share of samples allocated to
the most difficult arm.

One striking feature of the EVT algorithm seems to be the overall
iterations and budget allocated to arm \(V32\), which, as we have seen,
has a sample mean of zero. Something that might be hidden by this
aggregated view of all 5000 iterations are a share of iterations in
which the EVT completely fails by exploring wrong arms for long periods.
This effect might even be more drastic for the APT algorithm. Even
though it seems as if its behavior is very similar to SLR and EVT, the
overall performance is not better than the performance of uniform
sampling. Taking this reasoning into account, one can hypothesize that
the Bayes-UCB algorithm finds the most difficult arm uniformly across
most iterations as the rounds increase, while other algorithms either
find the most difficult arm right away or never and thus have many
iterations that end with a misclassification.

For the real data, the difference between Bayes-UCB and the other
algorithms is just as apparent. As is the effect of the time varying
means of the arms on the rate at which they are sampled. APT, SLR, and
EVT all focus quickly in many iterations on the still most difficult arm
\(V149\). However, in the case of the real data this behavior is not
rewarded in the end, as this leads to biased samples and thus wrong
estimation of the overall sample mean against which the algorithms are
evaluated. We see in the curves of arms \(V32\) and arm \(V154\) that
the EVT algorithm is less impacted by daily seasonalities, while APT
clearly is.

\includegraphics{Draft_files/figure-latex/unnamed-chunk-9-1.pdf}
